A sample computation

### 2024-05-27 draft_systoles.tex
family is everying

\DeclareMathOperator{\assign}{\coloneqq}        % Assingment operator
\newcommand{\R}{\mathbb{R}}        % Real
\newcommand{\Q}{\mathbb{Q}}        % rationals
\newcommand{\C}{\mathbb{C}}        % Complex
\renewcommand{\P}{\mathbb{P}}        % Projective
\newcommand{\A}{\mathbb{A}}        % Affine
\renewcommand{\O}{\mathcal{O}}        % Structure sheaf
\newcommand{\Id}{\mathbf{Id}}        % Identity
\newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
\newcommand{\GL}{\mathbf{GL}_3} % GL_3
\newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
\newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
\newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
\newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
\newcommand{\hence}{\Rightarrow}        % hence

@@

We can now define
\begin{equation}
\label{matrix_small_family}
A_{a, b} = 
\begin{pmatrix}
0 & 0 & b \\
1 & 0 & bp+a \\
0 & 1 & ap
\end{pmatrix}
\in \MatZ,
\end{equation}
\begin{equation}
\label{matrix_big_family}
\tilde A_{a, b} = \Id + p A_{a, b} =
\begin{pmatrix}
1 & 0 & b p \\
p & 1 & b p^2 + a p \\
0 & p & 1 + a p^2
\end{pmatrix}
\in \SLp.
\end{equation}
Here $\tilde A_{a, b}$ is indeed invertible thanks to 
\begin{multline}
	\label{computation_char_poly_of_a_tilde_family}
	\det({{\tilde{A}}_{a,b}})
= p^3 \det \left(\frac{1}p \Id + A_{a,b}\right) 
= - p^3 \det \left(\frac{-1}p \Id - A_{a,b}\right) \\
= - p^3 f_{a,b}\left(\frac{-1}p\right)
= 1 + a p^2 - (b p + a) p^2 + b p^3
= 1
.\end{multline}
Actually, almost all of the matrices from $\SLp$ are similar to $\tilde A_{p, q}$.
\begin{Prop}
Any matrix $B \in \SLp$ whose minimal polynomial is its characteristic polynomial satisfies
\[
B = C \tilde A_{a, b} C^{-1} 
,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
\end{Prop}
\begin{proof}
Pick $C \in \GL( \mathbb{Q} ) $ such that 
$$
C^{-1} \frac{1}{p}(B - \Id) C =
\begin{pmatrix}
0 & 0 & r_0 \\
1 & 0 & r_1 \\
0 & 1 & r_2
\end{pmatrix} \in 
\mathbf{SL}_3(\mathbb{Q})
.$$ 
This is possible to do because of the condition on $B$ and the theory of Frobenius normal form over $ \mathbb{Q}$ from linear algebra.

This matrix is actually integer $ C^{-1} \frac{1}{p}(B - \Id) C \in \MatZ$.
Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
Therefore, $ C^{-1} B C \in \SLp $.

It is left to show that 
$ C^{-1} B C = \tilde A_{a, b}$.
This follows from the equation $\det C^{-1} B C = 1$ which reads
\[
1 + r_2 p - r_1 p^2 + r_0 p^3 = 1
,\] 
which implies
 \[
r_0 = b, r_1 = bp + a, r_2 = ap,
\] 
for some $a, b \in \mathbb{Z}$.
\end{proof}

\section{The Unknown}
\begin{Prob}
What is the behavior of $\Reg, \deg $ for an arbitrary member of the family $\tilde A_{a, b}$?
\end{Prob}
Currently I don't really understand how to treat ramification.

>>>>

\DeclareMathOperator{\assign}{\coloneqq}        % Assingment operator
\newcommand{\R}{\mathbb{R}}        % Real
\newcommand{\Q}{\mathbb{Q}}        % rationals
\newcommand{\C}{\mathbb{C}}        % Complex
\renewcommand{\P}{\mathbb{P}}        % Projective
\newcommand{\A}{\mathbb{A}}        % Affine
\renewcommand{\O}{\mathcal{O}}        % Structure sheaf
\newcommand{\Id}{\mathbf{Id}}        % Identity
\newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
\newcommand{\GL}{\mathbf{GL}_3} % GL_3
\newcommand{\PGL}{\mathbf{GL}_3} % PGL_3
\newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
\newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
\newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
\newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
\newcommand{\hence}{\Rightarrow}        % hence
\newcommand{\minpoly}{\mathrm{minpoly}}        % minimal polynomial

@@

We can now define
\begin{equation}
\label{matrix_small_family}
A_{a, b} = 
\begin{pmatrix}
0 & 0 & b \\
1 & 0 & bp+a \\
0 & 1 & ap
\end{pmatrix}
\in \MatZ,
\end{equation}
\begin{equation}
\label{matrix_big_family}
\tilde A_{a, b} = \Id + p A_{a, b} =
\begin{pmatrix}
1 & 0 & b p \\
p & 1 & b p^2 + a p \\
0 & p & 1 + a p^2
\end{pmatrix}
\in \SLp.
\end{equation}
Here $\tilde A_{a, b}$ is indeed invertible thanks to 
\begin{multline}
	\label{computation_char_poly_of_a_tilde_family}
	\det({{\tilde{A}}_{a,b}})
= p^3 \det \left(\frac{1}p \Id + A_{a,b}\right) 
= - p^3 \det \left(\frac{-1}p \Id - A_{a,b}\right) \\
= - p^3 f_{a,b}\left(\frac{-1}p\right)
= 1 + a p^2 - (b p + a) p^2 + b p^3
= 1
.\end{multline}
Actually, a generic element in $\SLp$ is conjugate to $\tilde A_{a, b}$.
\begin{Prop}
An element $\tilde B \in \SLp$ is either
\begin{enumerate}
	\item $\Id$;
	\item conjugate to 
		\[
		\begin{pmatrix}
		1 & p & 0 \\
		0 & 1 & 0 \\
		0 & 0 & 1
		\end{pmatrix}
		\] 
		over $\mathbb{Q}$;
	\item conjugate to $\tilde A_{a, b}$ over $\mathbb{Q}$ for some $a, b \in \mathbb{Z}$.
\end{enumerate}
% That is, for any $B \neq \Id \in \SLp$ we have
% \[
% B = C \tilde A_{a, b} C^{-1} 
% ,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
% Moreover, $a, b$ are uniquely defined and $C$ is unique as an element of $\PGL( \mathbb{Q} )$.
\end{Prop}
The proof goes by applying the Frobenius normal form over $ \mathbb{Q}$.
\begin{proof}
Define $B = \frac{1}{p}(\tilde B - \Id) \in \MatZ$.

There are three cases:
the minimal polynomial $\minpoly_B$ of $B$ is of degree $1, 2$ or $3$.
\begin{enumerate}
	\item $\deg \minpoly_B = 1$.  Hence $\tilde B$ is scalar, hence $\Id$;
	\item $\deg \minpoly_B = 2$.  Hence $ \frac{\chi_B}{\minpoly_B}$ is linear, thus dividing $\minpoly_B$; we obtain $\chi_B$ decomposes into a product of linear multiples over $ \mathbb{Z}$
	$$
	\chi_{B} (\lambda) = (\lambda - a)^2 (\lambda - b), a, b \in \mathbb{Z}.
	$$
	The condition $ - p^3 \chi_{B}( \frac{-1}{p}) = \chi_{\tilde B}(0) = 1$ now reads
	\[
	(1 + a p)^2 (1 + b p ) = 1
	,\] 
	thus $a, b = 0$ if $p > 2$ and $B$ is nilpotent.
	Therefore $\tilde B$ is conjugate to 
	\[
	\begin{pmatrix}
	1 & p & 0 \\
	0 & 1 & 0 \\
	0 & 0 & 1
	\end{pmatrix}
	.\] 
	\item
	The Frobenius normal form of 
	$B$ 
	% $\frac{1}{p}(B - \Id)$ 
	consists of the unique $3 \times 3$ block.

	Pick $C \in \GL( \mathbb{Q} ) $ such that 
	$$
	C^{-1} B C =
	\begin{pmatrix}
	0 & 0 & r_0 \\
	1 & 0 & r_1 \\
	0 & 1 & r_2
	\end{pmatrix} \in 
	\mathbf{SL}_3(\mathbb{Q})
	.$$ 
	This matrix is actually integer $ C^{-1} B C \in \MatZ$.
	Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
	Therefore, $ C^{-1} \tilde B C \in \SLp $.

	We now show that 
	$ C^{-1} \tilde B C = \tilde A_{a, b}$.
	This follows from the equation $\det (C^{-1} \tilde B C) = 1$ which reads
	\[
	1 + r_2 p - r_1 p^2 + r_0 p^3 = 1
	,\] 
	implying
	 \[
	r_0 = b, r_1 = bp + a, r_2 = ap,
	\] 
	for some $a, b \in \mathbb{Z}$.
\end{enumerate}

\end{proof}

\section{The Unknown}
\begin{Prob}
What is the behavior of $\Reg, \deg $ for an arbitrary member of the family $\tilde A_{a, b}$?
\end{Prob}
